Brief Announcement: An Improved Lower Bound for Dynamic Balanced Graph Partitioning

### 2020-02-14 paper/paper.tex
general editing, partition => configuration

\usepackage[utf8]{inputenc}
%\usepackage{xspace}
\usepackage{balance}
\usepackage{amsmath,amssymb,amsfonts,mathtools,amsthm}
\usepackage[noend]{algpseudocode}
\usepackage{algorithm}

@@

So far, the competitive ratio of the best known algorithm was the component-based algorithm with the ratio $O(k^2\cdot \ell^2)$ \cite{repartition-disc}.
We present an optimal algorithm for $k=3$ in Section~\ref{sec:k3}.
In Section~\ref{sec:ppl}, we present a $O(k\cdot \ell)$-competitive algorithm for the restricted scenario
where the communication graph (i.e., request sequence) admits a \emph{perfect partition} (i.e., a multi-way cut of cost zero) that is also an optimal partition for the offline problem.
For this setting, the algorithm is optimal, as the lower bound utilized an input sequence that has a perfect partition.

@@

Formally, the online \emph{Balanced RePartitioning} problem (\OBRP{}) is defined as
follows. We are given a set of $n$ nodes,
initially arbitrarily partitioned into $\ell$~clusters,
each of size~$k$.
We say that nodes in $C \subset V$ are \emph{collocated} in a partition $P$
if they are assigned to the same cluster.
%\maciek{We can ommit the formal definition: "I.e.,
%$\forall u,v \in C: \mathit{cluster} (u,P) = \mathit{cluster} (v,P)$.". %However, if we use the notion of cluster later, then TODO: define cluster}
%\maciek{We can ommit the formal definition: "I.e.,
%$\forall u,v \in C: \mathit{cluster} (u,P) = \mathit{cluster} (v,P)$.". However, if we use the notion of cluster later, then TODO: define cluster}
The input consists of a sequence of pairwise communication requests
$\sigma = (u_1,v_1),$ $(u_2,v_2),$ $(u_3,v_3), \ldots$,
where a pair $(u_t,v_t)$ indicates that nodes $u_t$ and $v_t$ exchange a fixed amount of data.
Before serving the request,
the online algorithm
can re-partition,
i.e.,
it may (optionally) move some of the nodes into clusters different than their current assigned clusters.
Once the re-partition is executed,
we serve
the communication request between two  nodes at cost~0 if they are collocated,
or at cost~1 if they are located in different clusters.
The cost of migrating a node from one cluster to another
is~$\alpha \in \mathbb{Z}^+$.
For any algorithm $ALG$,
its cost,
denoted by $ALG(\sigma)$,
is the total cost of communications and
the cost of migrations performed by $ALG$ while serving the sequence $\sigma$.

@@

We assume OPT begins with the initial partition
$P_I = I_1, \dots, I_{\ell}$ and moves to the final partition
$P_F = F_1, \dots, F_{\ell}$.
 The \emph{distance} of a partition $P = C_1, \dots, C_{\ell}$ from the initial partition is the number of nodes in $P$ that do not reside in their initial cluster.
    That is,
    $\mathit{dist}(P, P_I) := \sum_{j=1}^{\ell} | C_j \setminus I_j |$. 
In other words,
at least $\mathit{dist}(P, P_I)/2$ node swaps are required in order to reach the partition $P$ from $P_I$.
Therefore,
$\OPT \geq \Delta:= dist(P_F, P_I) $.
The online algorithm PPL mimics OPT by minimizing the distance to the initial partition $P_I$ with each re-partitioning.
As a result,
ON never ends up with a partition that is more than $\Delta$ away from $P_I$.
This invariant ensures that ON does not pay too much while reaching $P_F$.

\begin{algorithm}
    \renewcommand{\algorithmicrequire}{\textbf{Input:}}
    \renewcommand{\algorithmicensure}{\textbf{Output:}}
    \begin{algorithmic}[1]
        \Require 
        $k, \ell$,
        initial partition $P_I$,
        sequence of  requests $\sigma_1, \dots, \sigma_N$ 
        \Ensure A final partition $P_F$ 
        \State For each node $v$ create a singleton component $C_v$ and add it to $\mathcal{C}$
        \State{$P_0 := P_I$}
         \label{line:initcomponents}
        \For {each  request $\sigma_t=\{u,v\}, 1 \leq t \leq N$}
        \State Let $C_1 \ni u$ and $C_2 \ni v$ be the container components
        \If{$C_1 \neq C_2$}
        \State Unite the two components into a single component $C'$ and
        $\mathcal{C} = (\mathcal{C}\setminus\set{C_1, C_2}) \cup ~\set{C'}$ \label{line:mergecomponents}
%        \If{$\mathit{cluster}(C_1, P_{t-1}) \neq \mathit{cluster}(C_2, P_{t-1})$}
        \If{$\mathit{cluster}(C_1, P_{t-1}) \neq \mathit{cluster}(C_2, P_{t-1})$}
        \Comment{if not in the same cluster}
        \State $P_{t} = \mathit{re\mhyphen partition}(P_{t-1}, P_I, \mathcal{C}$) \label{line:rebalance} 
        \Comment{move to $P$ closest to $P_I$}
        \EndIf
        \EndIf
        \EndFor
    \end{algorithmic}
    \caption{Perfect Partition Learner (PPL)}
    \label{alg:ppl}
      \end{algorithm}

%      \maciek{``re-partition'' procedure name should indicate the fact that this is a specific repartition that is close to initial configuration}
      We note that the $\mathit{re\mhyphen partition}$ at Line \ref{line:rebalance} replaces the current partition $P$ with a (perfect) partition closest to $P_I$.
Hence it never moves to a partition beyond distance $\Delta$.      
\begin{property} \label{prop:dist<OPT}
    Let $P$ be any partition chosen by Algorithm \ref{alg:ppl} at Line $\ref{line:rebalance}$.
    Then, $\mathit{dist}(P,P_I) \leq \Delta$.
\end{property}

@@

\begin{theorem}	\label{thm:upperbound}
    PPL reaches the final partition $P_F$ and it is $(2\cdot k\cdot\ell)$-competitive.
\end{theorem}
\begin{proof}
      On each inter-cluster request,
     the algorithm enumerates all $\ell$-way partitions of components
     that are in the same (closest) distance of $P_I$.
     That is, 
     once it reaches $\Delta = \mathit{dist} (P_{t}, P_I)$,
     it does not move to a partition
     $P_{t'}$, for any $t'>t,\mathit{dist} (P_{t'}, P_I) > \Delta$,
     before it enumerates all partitions at distance $\Delta$.
     Therefore,
     PPL eventually reaches $\Delta=\OPT$ and the partition $P_F$.
%    including the request that completes revealing of all components that are collocated in $P_F$.
    There are at most $(k-1)\cdot\ell < k\cdot\ell $ calls   to $\mathit{re\mhyphen partition}$
     (i.e., the number of internal edges in $P_F$).
    By Lemma \ref{lemma:rebalancecost},
    each re-partition costs at most $2\cdot\OPT$.
    The total cost is therefore at most $2\cdot\OPT\cdot k\cdot\ell$ which implies the competitive ratio.
%    \mahmoud{This is the cost of moving and the cost of remote comm. is not counted.
%    	So it is 4-competitive (?)}
 \end{proof}

@@

\begin{proof}
	We construct an instance of the problem with $\ell$ clusters 
	$\set{ S_1, S_2,\dots , S_{\ell}}, |S_i|  = k$.
	Let $I(C)$ denote the cluster where nodes in a component $C$ are located initially.
%	\maciek{So far components were not introduced; the best place to introduce these would be around the definition of \PPOBRP}.
	Fix any online algorithm \ALG{}.
	We construct the input sequence of requests for \ALG{} as follows.
	First,
	we issue $k-2$ (internal) requests so that \ALG{} can form a component of $k-1$
	nodes on the cluster $S_1$.
	Recall that internal requests does not incur any cost for \ALG{}.
	If \ALG{} at any point splits a component
	(i.e., spreads its nodes over two or more clusters),
	then we continue to issue requests between non-collocated nodes of the component until \ALG{} collocates the nodes of the component.
	If \ALG{} never collocates them then it is not competitive.

	Let $x_0$  be the only single node left on $S_1$ and  $y_0 \in S_2$ be any single node on $S_2$.
	Next,
	we issue the request between $x_0$ and $y_0$.
	Since this is an external request,
	\ALG{} joins them into one component and collocates them in some cluster other than $S_1$.
	For this,
	\ALG{} moves to a new configuration (i.e., partition)
	that replaces $x_0$ and $y_0$ with two other single nodes $x_1$ and $y_1$ respectively.

@@

>>>>

\usepackage[utf8]{inputenc}
%\usepackage{xspace}
\usepackage{balance}
\usepackage{amsmath,amsfonts,mathtools,amsthm}
\usepackage[noend]{algpseudocode}
\usepackage{algorithm}

@@

So far, the competitive ratio of the best known algorithm was the component-based algorithm with the ratio $O(k^2\cdot \ell^2)$ \cite{repartition-disc}.
We present an optimal algorithm for $k=3$ in Section~\ref{sec:k3}.
In Section~\ref{sec:ppl}, we present a $O(k\cdot \ell)$-competitive algorithm for the restricted scenario
where the communication graph (i.e., request sequence) admits a \emph{perfect partition} (i.e., a multi-way cut of cost zero) that is also an optimal configuration for the offline problem.
For this setting, the algorithm is optimal, as the lower bound utilized an input sequence that has a perfect partition.

@@

Formally, the online \emph{Balanced RePartitioning} problem (\OBRP{}) is defined as
follows. We are given a set of $n$ nodes,
initially arbitrarily partitioned into $\ell$~clusters,
each of size~$k$.
We say that nodes in $C \subset V$ are \emph{collocated}% in a partition $P$
if they are assigned to the same cluster.
%\maciek{We can ommit the formal definition: "I.e.,
%$\forall u,v \in C: \mathit{cluster} (u,P) = \mathit{cluster} (v,P)$.". %However, if we use the notion of cluster later, then TODO: define cluster}
%\maciek{We can ommit the formal definition: "I.e.,
%$\forall u,v \in C: \mathit{cluster} (u,P) = \mathit{cluster} (v,P)$.". However, if we use the notion of cluster later, then TODO: define cluster}
The input consists of a sequence of pairwise communication requests
$\sigma = (u_1,v_1),$ $(u_2,v_2),$ $(u_3,v_3), \ldots$,
where a pair $(u_t,v_t)$ indicates that nodes $u_t$ and $v_t$ exchange a fixed amount of data.
Before serving the request,
the online algorithm can re-partition,
i.e.,
it may (optionally) move some of the nodes into clusters different than their current assigned clusters.
Once the re-partition is executed,
we serve
the communication request between two  nodes at cost~0 if they are collocated,
or at cost~1 if they are located in different clusters.
The cost of migrating a node from one cluster to another
is~$\alpha \in \mathbb{Z}^+$.
For any algorithm $ALG$,
its cost,
denoted by $ALG(\sigma)$,
is the total cost of communications and
the cost of migrations performed by $ALG$ while serving the sequence $\sigma$.

@@

We assume OPT begins with the initial configuration
$P_I = I_1, \dots, I_{\ell}$ and moves to the final partition
$P_F = F_1, \dots, F_{\ell}$.
 The \emph{distance} of a configuration $P = C_1, \dots, C_{\ell}$ from the initial configuration is the number of nodes in $P$ that do not reside in their initial cluster.
    That is,
    $\mathit{dist}(P, P_I) := \sum_{j=1}^{\ell} | C_j \setminus I_j |$. 
In other words,
at least $\mathit{dist}(P, P_I)/2$ node swaps are required in order to reach the configuration $P$ from $P_I$.
Therefore,
$\OPT \geq \Delta:= dist(P_F, P_I) $.
 With each re-partitioning,
  PPL moves to a configuration that minimizes distance to the initial configuration $P_I$.
As a result,
ON never ends up in a configuration that is more than $\Delta$ away from $P_I$.
This invariant ensures that ON does not pay too much while reaching $P_F$.

\begin{algorithm}
    \renewcommand{\algorithmicrequire}{\textbf{Input:}}
    \renewcommand{\algorithmicensure}{\textbf{Output:}}
    \begin{algorithmic}[1]
        \Require 
        $k, \ell$,
        initial configuration $P_I$,
        sequence of  requests $\sigma_1, \dots, \sigma_N$ 
        \Ensure A final configuration $P_F$ 
        \State For each node $v$ create a singleton component $C_v$ and add it to $\mathcal{C}$
        \State{$P_0 := P_I$}
         \label{line:initcomponents}
        \For {each  request $\sigma_t=\{u,v\}, 1 \leq t \leq N$}
        \State Let $C_1 \ni u$ and $C_2 \ni v$ be the container components
        \If{$C_1 \neq C_2$}
        \State Unite the two components into a single component $C'$ and
        $\mathcal{C} = (\mathcal{C}\setminus\set{C_1, C_2}) \cup ~\set{C'}$ \label{line:mergecomponents}
%        \If{$\mathit{cluster}(C_1, P_{t-1}) \neq \mathit{cluster}(C_2, P_{t-1})$}
        \If{$\mathit{cluster}(C_1, P_{t-1}) \neq \mathit{cluster}(C_2, P_{t-1})$}
        \Comment{if not in the same cluster}
        \State $P_{t} = \mathit{re\mhyphen partition}(P_{t-1}, P_I, \mathcal{C}$) \label{line:rebalance} 
        \Comment{move to $P$ closest to $P_I$}
        \EndIf
        \EndIf
        \EndFor
    \end{algorithmic}
    \caption{Perfect Partition Learner (PPL)}
    \label{alg:ppl}
      \end{algorithm}

%      \maciek{``re-partition'' procedure name should indicate the fact that this is a specific repartition that is close to initial partition}
      We note that the $\mathit{re\mhyphen partition}$ at Line \ref{line:rebalance} replaces the current configuration $P$ with a (perfect) partition closest to $P_I$.
Hence it never moves to a configuration beyond distance $\Delta$.      
\begin{property} \label{prop:dist<OPT}
    Let $P$ be any configuration chosen by Algorithm \ref{alg:ppl} at Line $\ref{line:rebalance}$.
    Then, $\mathit{dist}(P,P_I) \leq \Delta$.
\end{property}

@@

\begin{theorem}	\label{thm:upperbound}
    PPL reaches the final configuration $P_F$ and it is $(2\cdot k\cdot\ell)$-competitive.
\end{theorem}
\begin{proof}
      On each inter-cluster request,
     the algorithm enumerates all $\ell$-way partitions of components
     that are in the same (closest) distance of $P_I$.
     That is, 
     once it reaches a configuration $P$ at distance $\Delta = \mathit{dist} (P, P_I)$,
     it does not move to a partition
     $P', \mathit{dist} (P', P_I) > \Delta$,
     before it enumerates all partitions at distance $\Delta$.
     Therefore,
     PPL eventually reaches $\Delta=\OPT$ and the configuration $P_F$.
%    including the request that completes revealing of all components that are collocated in $P_F$.
    There are at most $(k-1)\cdot\ell < k\cdot\ell $ calls   to $\mathit{re\mhyphen partition}$
     (i.e., the number of internal edges in $P_F$).
    By Lemma \ref{lemma:rebalancecost},
    each re-partition costs at most $2\cdot\OPT$.
    The total cost is therefore at most $2\cdot\OPT\cdot k\cdot\ell$ which implies the competitive ratio.
%    \mahmoud{This is the cost of moving and the cost of remote comm. is not counted.
%    	So it is 4-competitive (?)}
 \end{proof}

@@

\begin{proof}
	\mahmoud{mention $\alpha=1$}
	We construct an instance of the problem with $\ell$ clusters 
	$\set{ S_1, S_2,\dots , S_{\ell}}, |S_i|  = k$.
	Let $I(C)$ denote the cluster where nodes in a component $C$ are located initially.
%	\maciek{So far components were not introduced; the best place to introduce these would be around the definition of \PPOBRP}.
	Fix any online algorithm \ALG{}.
	We construct the input sequence of requests for \ALG{} as follows.
	First,
	we issue $k-2$ (internal) requests so that \ALG{} can form a component of $k-1$
	nodes on the cluster $S_1$.
	Recall that internal requests does not incur any cost for \ALG{}.
	If \ALG{} at any point splits a component
	(i.e., spreads its nodes over two or more clusters),
	then we continue to issue requests between non-collocated nodes of the component until \ALG{} collocates the nodes of the component.
	If \ALG{} never collocates them then it is not competitive.

	Let $x_0$  be the only single node left on $S_1$ and  $y_0 \in S_2$ be any single node on $S_2$.
	Next,
	we issue the request between $x_0$ and $y_0$.
	Since this is an external request,
	\ALG{} joins them into one component and collocates them in some cluster other than $S_1$.
	For this,
	\ALG{} moves to a new configuration
	that replaces $x_0$ and $y_0$ with two other single nodes $x_1$ and $y_1$ respectively.

@@

    We distinguish between three types of clusters: $C_1, C_2, C_3$,
    which we define as follows.
    A cluster of type $C_1$ the cluster contains $3$ singleton components (this is also the initial configuration of any cluster).
    A cluster of type $C_2$  contains one component of size $2$ and one component of size $1$.
    Finally, a cluster of type $C_3$  contains one component of size $3$.
